Data Specialization for a Staged Lambda Calculus

### 2014-11-12 paper/example.tex
finished (a draft of) section 2

Now, consider an application where we perform many order statistics queries on
the same collection \texttt{l}, but with $m$ different ranks \texttt{k}.
Certainly, it is possible to implement this with $m$ calls to
\texttt{quickselect}:
%
\begin{lstlisting}
quickselect l @$k_1$@
quickselect l @$k_2$@
 @$\vdots$@ 
quickselect l @$k_m$@.
\end{lstlisting}
%
Can we do better?

@@

An astute programmer might notice that much of the code in \texttt{quickselect}
does \emph{not} depend on the rank \texttt{k}---for example, the list is
partitioned before \texttt{k} is ever used. Moreover, while the recursive calls
to \texttt{qSelect} are guarded by a comparison to \texttt{k}, the lists passed
in to those recursive calls also don't depend on \texttt{k}. Thus, if we are
willing to change the evaluation order of the language (and evaluate under the
\texttt{case} statement), it should be possible to set things up so that
\emph{all} the computation involving \texttt{l} is performed before \emph{any}
of the computation involving \texttt{k}.

We can formalize this intuition by writing quickselect in \lang\
(\ref{sec:semantics}), a \emph{staged}, typed lambda calculus. In \lang, we can
directly express the idea that the argument \texttt{l} is known at the
\emph{first} stage of the computation, while the argument \texttt{k} is only
known at the \emph{second}, and all purely-stage-\bbone\ computation occurs
before any stage \bbtwo\ computation.

A staged version of quickselect is written in \ref{fig:qs-staged}, where stage
\bbone\ code is red, and stage \bbtwo\ code is blue. Whereas before
\texttt{qSelect} took a \textrm{list} and \rmint\ and returned an \rmint, now we
must be more careful---at stage \bbone, \texttt{qsStaged} takes a \textrm{list}
and a $\fut\rmint$ (an $\rmint$ \emph{in the future}), and returns a
$\fut\rmint$.

At stage \bbone, all terms of non-circle type are available for
immediate use. Indeed, we are able to case on \texttt{l} as usual, and in the
\texttt{Cons} branch, pass it to the stage \bbone\ function \texttt{partition}.

\crem{Very rough text for the remainder of the section. Be a bit more explicit.}

However, \texttt{k} has type $\fut\rmint$, so we cannot use it as part of a
stage \bbone\ computation. Luckily, we are trying to produce a $\fut\rmint$. To
create one of these, we can wrap $\next$ around a stage \bbtwo\ computation of
an integer; in that computation, we can wrap $\prev$ around the $\fut\rmint$
\texttt{k} in order to use it as a stage \bbtwo\ integer.

(There's a subtle difference here between an \emph{int later} at the first
stage, and an int at the second stage.)

Essentially, $\next$ allows us to pass around opaque stage \bbtwo\ computations
at stage \bbone, and the $\prev$ construct lets us unwrap those computations for
use at stage \bbtwo.

For convenience, the code also uses $\pause$ blocks. This construct has same
stage signature as $\prev$, but instead of ``unwrapping'' $\fut$ types it simply
promotes integers from stage \bbone\ to stage \bbtwo.%
\footnote{$\pause$ can be implemented from our other language features, with
some effort. We provide it as a primitive to shorten examples. Furthermore, it
would be wise to extend $\pause$ to other base types, if we had them, and to
products and sums thereof.  This is related to the notion of {\em mobility} in
\cite{murphy05} and {\em stability} in \cite{krishnaswami13}.}

The process of adding staging annotations to unstaged code has been the subject
of extensive research under the name of \emph{binding time analysis}. In this
paper, we assume that these annotations have been provided by the programmer (or
perhaps a binding time analysis tool), and do not consider the problem of
generating such annotations. There are many ways to annotate any program,
including \texttt{qselect}; the version presented here matches the intuition for
the maximal amount of work that can be performed at stage \bbone.

The purpose of writing quickselect in \lang\ is that, because \texttt{qsStaged}
typechecks, it is a valid way of dividing the program into two stages, where the
stage \bbone\ computations can all be performed prior to the stage \bbtwo\
computations. In \ref{sec:semantics} we provide an evaluation strategy for
\lang\ which validates that claim.

%The construct $\next$ occurs in a stage \bbone\ context and indicates that the
%contents of its block are stage \bbtwo, whereas $\prev$ occurs in a stage
%\bbtwo\ context and indicates that the contents of its block is stage \bbone.

@@

@@

Because we have found a valid staging \texttt{qsStaged}, one might ask what
happens if we split this function into \emph{two} functions, one of which
performs all the stage \bbone\ computations given \texttt{l}, and the second
which performs all the stage \bbtwo\ computations given \texttt{k} and the
results of the previous.

Intuitively, what should this be? All calls to \texttt{partition} are in stage
\bbone, so this means we build a binary search tree from \texttt{l}. All
decisions on which side to recur down happen in stage \bbtwo, so this means we
traverse the tree once we have the rank \texttt{k}.

(There are two kinds of things we must pass to the second function---which stage
\bbtwo\ computations to use, and also anything that was held.)

\crem{introduce boundary terminology}

Formally, we wish to derive two functions, \texttt{qSelect1} and
\texttt{qSelect2}, which respectively contain all the stage \bbone\ and stage
\bbtwo\ content of \texttt{qSelect}, and are moreover semantically equivalent to
\texttt{qSelect} in the sense that:
\\
\texttt{qSelect l k = qSelect2 (qSelect1 l) k}.
\\

Since \texttt{quickselect1} requires only on a first stage
argument \texttt{l}, we can run it once to produce a boundary data
structure that can be used for all subsequent queries on \texttt{l},
e.g., we can perform $m$ queries by simply running:
%
\begin{lstlisting}
let b = qSelect1 l in
  qSelect2 b @$k_1$@
  qSelect2 b @$k_2$@
   @$\vdots$@ 
  qSelect2 b @$k_m$@.
\end{lstlisting}

Indeed, in \ref{sec:splitting,sec:implementation} we describe an algorithm for
splitting staged terms, which on \texttt{qsStaged} actually emits the algorithm
in \ref{fig:qssplit}.

Assuming that the list~\texttt{l} contains $n$ elements, this optimization
changes the asymptotic complexity from expected (randomized) $\Theta(n \cdot m)$
to $\Theta(n\log{n} + m\log{n})$, which for any $m \approx n$ reduces the
complexity from $\Theta (n^2)$ to $\Theta(n\log{n})$---a near linear time
improvement.

%Note that the desired optimized code shows above is intellectually more
%sophisticated than the code that we have started with: the optimized code is
%able to create a data structure, a balanced binary tree augmented with indexing
%information, and use a binary search technique over this tree to compute the
%result asymptotically more efficiently.
%In fact, based our teaching experience, we can imagine this kind of problem to
%be a moderately difficult exam question in an undergraduate algorithms class, as
%it not only requires understanding of data structures such as binary search
%trees but also requires modifying them to augment with indexing information to
%support rank-based search.

>>>>

Now, consider an application where we perform many order statistics queries on
the same collection \texttt{l}, but with $m$ different ranks
$\mathtt{k_1},\dots,\mathtt{k_m}$.
Certainly, it is possible to implement this with $m$ calls to
\texttt{quickselect}:
%
\begin{lstlisting}
quickselect l @$\mathtt{k_1}$@
quickselect l @$\mathtt{k_2}$@
 @$\vdots$@ 
quickselect l @$\mathtt{k_m}$@.
\end{lstlisting}
%
Can we do better?

@@

An astute programmer might notice that much of the code in \texttt{quickselect}
does \emph{not} depend on the rank \texttt{k}; for example, the list is
partitioned before \texttt{k} is ever used. Moreover, while the recursive calls
to \texttt{qSelect} are guarded by a comparison to \texttt{k}, \texttt{k} does
not determine the arguments to those recursive calls---it only determines which
call is made. Thus, if we are willing to change the evaluation order of the
language (and evaluate under the \texttt{case} statement), it should be possible
to set things up so that \emph{all} the computation involving \texttt{l} is
performed before \emph{any} of the computation involving \texttt{k}.

We can formalize this intuition by writing quickselect in a \emph{staged}
language. In this paper, we choose \lang\ (\ref{sec:semantics}), a staged, typed
lambda calculus. This allows us to directly express the idea that the argument
\texttt{l} is known at the \emph{first} stage of the computation, while the
argument \texttt{k} is only known at the \emph{second}, and all
purely-stage-\bbone\ computation occurs before any stage \bbtwo\ computation.

We define a staged version of quickselect in \ref{fig:qs-staged}, writing stage
\bbone\ code in red, and stage \bbtwo\ code in blue. Whereas \texttt{qSelect}
sends a \textrm{list} and \rmint\ to an \rmint, \texttt{qsStaged} is more
precise---it is a stage \bbone\ function which takes a \textrm{list} and a
$\fut\rmint$ (an $\rmint$ \emph{in the future}), and returns a $\fut\rmint$.

The stage \bbone\ portions of \texttt{qsStaged} look like \texttt{qSelect}: any
term of non-circle type is available for immediate use. Indeed, we case on
\texttt{l} as usual, and in the \texttt{Cons} branch, \texttt{partition} it.

The next part of the original \texttt{qSelect} function depends on \texttt{k},
which has type $\fut\rmint$, indicating that it is only available to stage
\bbtwo\ computations. Since \texttt{qsStaged} itself produces a $\fut\rmint$,
the remainder of the function is stage \bbtwo\ code.

The $\next$ wraps a stage \bbtwo\ expression of type $\rmint$ (the \texttt{case}
expression) as a stage \bbone\ expression of type $\fut\rmint$ (the result type
of \texttt{qsStaged}). Inside the \texttt{LT} and \texttt{GT} branches, the
$\prev$ unwraps stage \bbone\ $\fut\rmint$s (the results of the recursive calls
to \texttt{qsStaged}) as stage \bbtwo\ $\rmint$s. In the \texttt{EQ} branch,
$\pause$ promotes an integer at stage \bbone\ (the head of \texttt{l}) directly
to an integer at stage \bbtwo.%
\footnote{$\pause$ can be implemented from our other language features, with
some effort. We provide it as a primitive to shorten examples. Furthermore, it
would be wise to extend $\pause$ to other base types, if we had them, and to
products and sums thereof.  This is related to the notion of {\em mobility} in
\cite{murphy05} and {\em stability} in \cite{krishnaswami13}.}

The process of adding staging annotations ($\fut$ types, $\next$, $\prev$, and
$\pause$) to unstaged code has been the subject of extensive research under the
name of \emph{binding time analysis}. In this paper, we assume that these
annotations have been provided by the programmer (or perhaps a binding time
analysis tool), and do not consider the problem of generating such annotations.
There are many ways to annotate any program, including \texttt{qSelect}; we
chose annotations which maximize the work performed at stage \bbone.

%The purpose of writing quickselect in \lang\ is that, because \texttt{qsStaged}
%typechecks, it is a valid way of dividing the program into two stages, where the
%stage \bbone\ computations can all be performed prior to the stage \bbtwo\
%computations. In \ref{sec:semantics} we provide an evaluation strategy for
%\lang\ which validates that claim.

@@

%pass around opaque stage \bbtwo\ computations at stage \bbone

@@

An astute programmer, having noticed that \texttt{quickselect} can be staged in
this fashion, might try to split it into a pair of functions, one which performs
all the work depending only on \texttt{l} (stage \bbone), and one which uses
\texttt{k} and the result of that work on \texttt{l} to compute the element with
rank \texttt{k} in \texttt{l}.

Intuitively, \texttt{l} determines the result of all calls to
\texttt{partition}, and \texttt{k} only determines which calls are made. So we
can preprocess \texttt{l} by recursively dividing it into halves smaller and
greater than the pivot---that is, building a binary search tree. Then, once we
have \texttt{k}, we can recur on this tree, choosing whichever branch has the
\texttt{k}${}^\textit{th}$ leftmost element until we reach a leaf. Crucially,
because \texttt{partition} has no stage \bbtwo\ content, we need not perform any
of its work at stage \bbtwo!

We have implemented this splitting of quickselect in \ref{fig:qs-split}.
\texttt{qSelect1} builds a binary search tree from the list \texttt{l}, and
\texttt{qSelect2} takes such a tree and a rank \texttt{k} and computes the
answer. This allows us to efficiently perform many order statistics queries on
\texttt{l} by caching the tree and reusing it for many different ranks
$\mathtt{k_1},\dots,\mathtt{k_m}$:
%
\begin{lstlisting}
let b = qSelect1 l in
  qSelect2 b @$k_1$@
  qSelect2 b @$k_2$@
   @$\vdots$@ 
  qSelect2 b @$k_m$@.
\end{lstlisting}

Assuming \texttt{l} contains $n$ elements, this optimization changes the
asymptotic complexity from expected (randomized) $\Theta(n \cdot m)$ to
$\Theta(n\log{n} + m\log{n})$, which for any $m \approx n$ reduces the
complexity from $\Theta (n^2)$ to $\Theta(n\log{n})$---a near linear time
improvement.

In this paper, we develop a splitting algorithm
(\ref{sec:splitting,sec:implementation}) which, for any term $e$ in \lang,
produces a pair of terms $e_1,e_2$ containing respectively all the stage \bbone\
and stage \bbtwo\ content of $e$, such that $e_2~e_1$ is equivalent to $e$. In
the case of \texttt{qsStaged}, our splitting algorithm produces the algorithm
described above.

%\texttt{qSelect l k = qSelect2 (qSelect1 l) k}.

%Because the tree passes information across the stage boundary, we call it the
%\emph{boundary data structure}.

%Note that the desired optimized code shows above is intellectually more
%sophisticated than the code that we have started with: the optimized code is
%able to create a data structure, a balanced binary tree augmented with indexing
%information, and use a binary search technique over this tree to compute the
%result asymptotically more efficiently.

%In fact, based our teaching experience, we can imagine this kind of problem to
%be a moderately difficult exam question in an undergraduate algorithms class, as
%it not only requires understanding of data structures such as binary search
%trees but also requires modifying them to augment with indexing information to
%support rank-based search.
